Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-09 ms.tex

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung radiation in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated to these
temperatures primarily through collisionless structure-formation shocks that form as a result of
the hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles
\citep[see, e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do
indeed feature mega-parsec scale halos of non-thermal radio emission, indicative of a population of
relativistic electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. Recent observations of possible non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the
basis of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution
in the Coma cluster with the VERITAS and \emph{Fermi}-LAT upper limits from this work. This
approach includes (1) an analytic model derived from cosmological hydrodynamical simulations of the
formation of galaxy clusters, (2) a simplified multi-frequency analytical approach that assumes a
constant CR-to-thermal energy density and a power-law spectrum in momentum, and (3) an approach
that uses the observed luminosity of the giant radio halo in Coma to place a lower limit on the
expected $\gamma$-ray flux in the hadronic model -- independent of the magnetic field distribution.
This translates into a minimum CR pressure which, if challenged by tight gamma-ray
limits/detections, enables scrutinizing the hadronic interaction model of the formation of giant
radio halos. Modelling the CR distribution through different techniques enable to bracket our
ignorance about the underlying plasma physics that shape the CR distribution and hence reflect the
Bayesian priors that are imposed on the modelling
\citep[see][for a discussion]{article:PinzkePfrommerBergstrom}.

@@

\subsection{Simulation-based approach}
We adopt the universal spectral and spatial gamma-ray model developed by
\citet{article:PinzkePfrommer:2010} to estimate the emission from decaying neutral pions that
dominates over the IC emission from primary and secondary electrons above 100 MeV in clusters.
Given a density profile as, e.g., inferred by cosmological simulations or X-ray observations, the
analytic approach models the CR distribution and the associated radiative emission processes from
radio to the gamma-ray band. This formalism was derived from high-resolution simulations of
clusters of galaxies that included radiative hydrodynamics, star formation, supernova feedback, and
followed CR physics by tracing the most important injection and loss processes self-consistently
while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin, article:MiniatiRyuKangJones:2001,
article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
article:Pfrommer:2008}.

The overall normalization of the CR and gamma-ray distribution scales with the maximum acceleration
efficiency at structure formation shock waves. Following recent observations at supernova remnants
\citep{article:Helder_etal:2009} as well as theoretical studies \citep{article:KangJones:2005}, we
adopt an optimistic but nevertheless realistic value of this parameter and assume that 50\% of the
dissipated energy at strong shocks is injected into CRs while this efficiency rapidly decreases for
weaker shocks. Since the vast majority of internal formation shocks (merger and flow shocks) are
weak shocks with Mach numbers $M\lesssim3$ \citep[e.g.,][]{article:Ryu_etal:2003}, they do not
significantly contribute to the CR population in clusters. Instead, strong shocks during the
formation epoch of clusters and strong accretion shocks at the present time (at the boundary of
voids and filaments/super-cluster regions) dominate the injection of CRs which are adiabatically
transported through the cluster. Hence, these model predictions provide a plausible upper limit for
the CR contribution from structure formation shocks in galaxy clusters which can be scaled with the
effective acceleration efficiency. Other possible CR sources, such as AGN and starburst-driven
galactic winds have been neglected for simplicity which could in principle increase the expected
gamma-ray yield.

@@

We adopt the electron density profile for the Coma cluster that has been inferred from ROSAT X-ray
observations \citep{article:BrielHenryBohringer:1992} and model the spatial and spectral
distribution of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the
contribution of cluster galaxies. Figure \ref{fig:spectrum} shows the expected spectrum of Coma,
normalized to the total flux within the virial radius (dotted lines) and the angular apertures
tested in this study (solid lines). The gamma-ray flux limit of {\em Fermi} in the energy interval
of 1--3 GeV ($<0.4^\circ$) is most constraining. However, since the model flux within the finite
energy range of 1--3 GeV is smaller by a factor of 1.5 in comparison to the total flux above 1 GeV,
it falls below the flux upper limits which do not yet constrain the simulation model. In our
further analysis, we use the most constraining {\em Fermi} flux limits in the energy interval of
1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above 220 GeV.

Figure \ref{fig:xcr} shows the relative CR pressure,
$X_{\CR} = \expval{P_{\CR}}/\expval{P_\rmn{th}}$ as a function of radial distance, $R$, from the
Coma cluster center (red lines) and contained within $R$ (blue lines).  We normalize the radii by
the virial radius, $R_\rmn{vir}$, that we define as the radius of a sphere enclosing a mean density
that is 200 times the critical density of the Universe.  To compute the CR pressure, we assume a
low-momentum cutoff of the CR distribution at $q = 0.8\,m_p c$ as found in the cosmological
simulations which reflects the high cooling rates at low CR energies. The relative CR pressure
rises toward the outer regions on account of the higher efficiency of CR acceleration at the
peripheral strong accretion shocks compared to weak central flow shocks. Adiabatic compression of a
mixture of CRs and thermal gas disfavours the CR pressure relative to the thermal pressure on
account of the softer equation of state of CRs. The weak increase of $X_{\CR}$ toward the core is
due to the comparably fast thermal cooling of gas which diminishes the thermal support relative to
that in CRs.

Table \ref{table:constraints} shows the expected gamma-ray fluxes within the regions of interest.
In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), those are a factor of 7.2 below the inferred upper limits of VERITAS (see Table
\ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches the simulated one
at energies $E_\gamma=200$ GeV). Hence we scale the averaged relative CR pressure of our model by
that factor, keep the spatial behavior and obtain an integrated limit of the relative CR pressure
of $X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit within the cluster virial radius
of $X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}). Equivalently, the most constraining
{\em Fermi} upper limit in the energy interval of 1--3 GeV ($<0.4^\circ$) is a factor of 1.1 larger
than our model predictions (assuming $\alpha=2.1$ which is very close to the simulated spectral
index for the energy range of 1-3 GeV). Scaling our integrated relative CR pressure profile yields
the constraint of $X_{\CR}<0.018$ within 0.4$^{\circ}$ that translates to a limit within the
cluster virial radius of $X_\CR<0.024$ (dashed lines of Figure \ref{fig:xcr}).

@@

\subsection{Simplified analytical model}
To complement the simulation-inspired analysis and to facilitate comparison with earlier analytical
work, we repeat the data analysis with a simplified analytical model that assumes a power-law CR
spectrum and a constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs}. To be
independent of additional assumptions and in line with earlier work in the literature, we do not
impose a low-momentum cutoff on the CR distribution function, i.e., we adopt $q=0$. Since a priori,
the CR spectral index is unconstrained, we vary it within plausible ranges of
$2.1<\alpha_{\rmn{p}}<2.5$. This spectral range is compatible with the radio spectral index of the
giant radio halo of Coma after accounting for the spectral steepening at frequencies
$\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}\footnote{Assuming a magnetic field of 1 $\mu$G, the CR
protons responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are
$\sim$ 20 times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.}.
Table \ref{table:constraints} shows the resulting constraints on $X_\CR$. The VERITAS flux upper
limits strongly depend on $\alpha$ due to the comparably large lever arm from GeV energies (that
dominate the CR pressure) to TeV-energies. The limits within 0.2$^{\circ}$ are most constraining
yielding limits on $X_\CR$ between 0.048 and 0.43 (for $\alpha$ varying between 2.1 and 2.5), with
a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close to the spectral index predicted by the
simulations around 220 GeV). This limit is more constraining by 50\% in comparison to the model
derived from simulations, which gives $X_\CR<0.15$. This difference is explained by the concave
curvature of the simulated spectrum which accumulates additional pressure toward GeV energies in
comparison to a pure power-law spectrum of $\alpha=2.3$.

The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an aperture of
0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only slightly
worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.  This is comparable to the $X_\CR$-constraint
in our simulation-based model of $X_\CR<0.024$ (evaluated within the cluster virial radius).
Naturally, with the Fermi limits we probe the region around GeV-energies that dominate the CR
pressure and we do not expect any differences in comparison to the simulation-based models on
account of the concave curvature of the universal CR spectrum found in the simulations.

\subsection{Minimum gamma-ray flux}
For clusters that host radio halos we are able to derive a minimum gamma-ray flux in the hadronic
model of CR interactions. The idea is based on the fact that a stationary distribution of CR
electrons loses all its energy to synchrotron radiation for strong magnetic fields
($B \gg B_{\rmn{CMB}} \simeq 3.2 \mu$G) so that the ratio of gamma-ray to synchrotron flux becomes
independent of the spatial distribution of CRs and thermal gas \citep{article:Voelk:1989,
article:Pohl:1994, article:Pfrommer:2008}, in particular with $\alpha_{\nu}\simeq 1$ as the
observed synchrotron spectral index.  Hence we can derive a minimum gamma-ray flux in the hadronic
model
\begin{equation}
\label{eq:Fmin}
\mathcal{F}_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\section{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous section that we can obtain an absolute lower limit on the gamma-ray
emission in the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. Hence, we can
turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR
pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio
emission which, however, assumes the validity of the hadronic model
\citep{article:PfrommerEnsslin:2004a}. Lowering the gamma-ray constraint will tighten (increase)
the magnetic field limit. Once this gets in conflict with magnetic field measurements by means of
other methods, e.g., Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the
magnetic-field strength by, e.g., background sources through clusters are degenerate with the
magnetic coherence scale and may be biased by the unknown correlation between magnetic and density
fluctuations.}, this would challenge the hadronic model of radio halos. The method that we use to
constrain the magnetic field inherits a dependence on the assumed spatial structure that we
parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact,
the magnetic field in the Coma cluster is among the best constrained, because its proximity permits
RM observations of seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the
central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary
(within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to
address this dependence. We proceed as follows:

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits (this work, Tables~\ref{table:results} and \ref{table:fermi})
for different integration radii (shown with different colors), assuming $\alpha=2.1$. These are
compared to simulated integrated spectra of VHE gamma-ray emission, $E_\gamma F_\gamma(>E_\gamma)$,
from the Coma cluster (again within different apertures). Note that the {\em Fermi} limit for the
energy interval of 1--3 GeV within the aperture of 0.4$^\circ$ is most constraining. Since the
model flux within the finite energy range of 1--3 GeV is smaller by a factor of 1.5 in comparison
to the total flux above 1 GeV, it falls below the flux upper limits which do not yet constrain the
simulation model.}
\label{fig:spectrum}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f5.eps}}
\end{center}
\caption{CR-to-thermal pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$, as a
function of radial distance from the center of the Coma. Model predictions are shown by dashed
lines \citep{article:PinzkePfrommer:2010} while these are scaled to match the most constraining
VERITAS upper limits within 0.2$^\circ$ (solid) and {\em Fermi} upper limits within 0.4$^\circ$
(dashed). We compare differential $X_\CR$-profiles (red) to integrated profiles
$X_{\CR}(<R/R_{\rmn{vir}})=\int_0^R P_\CR\, dV / \int_0^R P_{\rmn{th}}\, dV$ (blue) which we use to
compare to the upper limits.}
\label{fig:xcr}
\end{figure*}

>>>>

The ICM is a hot, $T\sim 10^{8}$ K, plasma emitting thermal bremsstrahlung radiation in the soft
X-rays regime \citep[see, e.g.,][]{article:Petrosian:2001}. This plasma has been heated to these
temperatures primarily through collisionless structure-formation shocks that form as a result of the
hierarchical merging and accretion processes. Such shocks and turbulence in the ICM gas and
intra-cluster magnetic fields also provide a means to efficiently accelerate particles \citep[see,
e.g.,][]{article:ColafrancescoBlasi:1998, article:Ryu_etal:2003}. Many clusters do indeed feature
mega-parsec scale halos of non-thermal radio emission, indicative of a population of relativistic
electrons and positrons (leptons) and magnetic fields permeating the ICM
\citep{article:Cassano_etal:2010}. The two competing theories seeking to explain the radio halo
emission feature the ``hadronic model'' where the radio-emitting electrons are produced in inelastic
collisions of CR ions with thermal gas protons from the ICM \citep[][and references
therein]{article:EnsslinPfrommerMiniatiSubramanian:2011} and the ``re-acceleration model'' where a
CR electron population within the ICM at energies around 100 MeV might be sufficiently long lived so
that during states of powerful ICM turbulence, e.g., after a cluster merger, non-resonant
interactions of these CR electrons with plasma waves might be efficient enough to accelerate them to
high enough energies ($\sim 10$~GeV) so that they emit observable radio emission \citep[][and
references therein]{2010arXiv1008.0184B}.  Recent observations of possible non-thermal emission from
clusters in the the extreme ultraviolet \citep[EUV; ][]{article:SarazinLieu:1998} and hard X-rays
\citep{article:RephaeliGruber:2002, article:Fusco-Femiano_etal:2004, article:Eckert_etal:2007} may
provide further indications of relativistic particle populations in clusters, although the
interpretation of these observations as non-thermal diffuse emission have been disputed on the basis
of more sensitive observations \citep[see, e.g.,][]{article:Ajello_etal:2009,
  article:Ajello_etal:2010, article:Wik_etal:2009}.

@@

\section{Gamma Ray Emission from Cosmic Rays}
We decided to adopt a multi-faceted approach to constrain the CR-to-thermal pressure distribution in
the Coma cluster with the VERITAS and \emph{Fermi}-LAT upper limits from this work. This approach
includes (1) a simplified multi-frequency analytical approach that assumes a constant CR-to-thermal
energy density and a power-law spectrum in momentum, (2) an analytic model derived from cosmological
hydrodynamical simulations of the formation of galaxy clusters, and (3) an approach that uses the
observed luminosity of the giant radio halo in Coma to place a lower limit on the expected
$\gamma$-ray flux in the hadronic model -- where the radio-emitting electrons are secondaries from
CR interactions and which is independent of the magnetic field distribution.  This translates into a
minimum CR pressure which, if challenged by tight gamma-ray limits/detections, enables scrutinizing
the hadronic interaction model of the formation of giant radio halos. Alternatively, by realizing a
spatial CR distribution that is just consitent with the flux upper limits and requiring the model to
match the observed radio data, this enables us to derive a lower limit on the magnetic field
distribution. This approach, however, assumes the validity of the hadronic interaction model in
explaining the radio halo data.  Modelling the CR distribution through different techniques enable
us to bracket our ignorance about the underlying plasma physics that shape the CR distribution and
hence reflect the Bayesian priors that are imposed on the modelling \citep[see][for a
discussion]{article:PinzkePfrommerBergstrom}.

@@

\subsection{Simplified analytical model}
\label{sec:simple}

We start by adopting a simplified analytical model that assumes a power-law CR spectrum and a
constant relative CR pressure, i.e., we adopt the {\em isobaric model of CRs} following the approach
of \citet{article:PfrommerEnsslin:2004b}. To be independent of additional assumptions and in line
with earlier work in the literature, we do not impose a low-momentum cutoff on the CR distribution
function, i.e., we adopt $q=0$. Since a priori, the CR spectral index is unconstrained, we vary it
within plausible ranges of $2.1<\alpha_{\rmn{p}}<2.5$. This spectral range is compatible with the
radio spectral index of the giant radio halo of Coma after accounting for the spectral steepening at
frequencies $\nu\sim5~\rmn{GHz}$ due to the Sunyaev-Zel'dovich effect
\citep{article:PfrommerEnsslin:2004b}\footnote{Assuming a magnetic field of 1 $\mu$G, the CR protons
  responsible for the GHz radio emitting electrons have an energy of $\sim100$ GeV and are $\sim$ 20
  times less energetic than those CR protons responsible for 200-GeV gamma-ray emission.}.  To model
the thermal pressure, we adopt the electron density profile for the Coma cluster that has been
inferred from ROSAT X-ray observations \citep{article:BrielHenryBohringer:1992} and use a constant
temperature of $kT= 8.25$~keV throughout the virial region.

Table \ref{table:constraints} shows the resulting constraints on the relative CR pressure, $X_{\CR}
= \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius, $R_\rmn{vir}=2.2$~Mpc,
that we define as the radius of a sphere enclosing a mean density that is 200 times the critical
density of the Universe. The VERITAS flux upper limits strongly depend on $\alpha$ due to the
comparably large lever arm from GeV energies (that dominate the CR pressure) to TeV-energies. The
limits within 0.2$^{\circ}$ are most constraining yielding limits on $X_\CR$ between 0.048 and 0.43
(for $\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close
to the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV).  The {\em Fermi} limits depend only weakly on $\alpha$ and are most constraining for an
aperture of 0.4$^{\circ}$.  The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$ and only
slightly worse for $\alpha\simeq 2.1$ yielding $X_\CR<0.025$.

\subsection{Simulation-based approach}

We complement the simplified analytical analysis with a more realistic and predictive approach
derived from cosmological hydrodynamical simulations. We adopt the universal spectral and spatial
gamma-ray model developed by \citet{article:PinzkePfrommer:2010} to estimate the emission from
decaying neutral pions that dominates over the IC emission from primary and secondary electrons
above 100 MeV in clusters.  Given a density profile as, e.g., inferred by cosmological simulations
or X-ray observations, the analytic approach models the CR distribution and the associated radiative
emission processes from radio to the gamma-ray band. This formalism was derived from high-resolution
simulations of clusters of galaxies that included radiative hydrodynamics, star formation, supernova
feedback, and followed CR physics by tracing the most important injection and loss processes
self-consistently while accounting for the CR pressure in the equation of motion
\citep{article:PfrommerSpringelEnsslinJubelgas, article:EnsslinPfrommerSpringelJubelgas:2007,
  article:JubelgasSpringelEnsslinPfrommer:2008}. The results are in line with earlier numerical
results on some of the overall characteristics of the CR distribution and the associated radiative
emission processes \citep{article:DolagEnsslin, article:MiniatiRyuKangJones:2001,
  article:Miniati:2003, article:Pfrommer_etal:2007, article:PfrommerEnsslinSpringel:2008,
  article:Pfrommer:2008}.

The overall normalization of the CR and gamma-ray distribution scales (non-linearly) with the
maximum acceleration efficiency at structure formation shock waves. Following recent observations at
supernova remnants \citep{article:Helder_etal:2009} as well as theoretical studies
\citep{article:KangJones:2005}, we adopt an optimistic but nevertheless realistic value of this
parameter and assume that 50\% of the dissipated energy at strong shocks is injected into CRs while
this efficiency rapidly decreases for weaker shocks. Since the vast majority of internal formation
shocks (merger and flow shocks) are weak shocks with Mach numbers $M\lesssim3$
\citep[e.g.,][]{article:Ryu_etal:2003}, they do not significantly contribute to the CR population in
clusters. Instead, strong shocks during the formation epoch of clusters and strong accretion shocks
at the present time (at the boundary of voids and filaments/super-cluster regions) dominate the
injection of CRs which are adiabatically transported through the cluster. Hence, these model
predictions provide a plausible upper limit for the CR contribution from structure formation shocks
in galaxy clusters which can be scaled with the effective acceleration efficiency. Other possible CR
sources, such as AGN and starburst-driven galactic winds have been neglected for simplicity which
could in principle increase the expected gamma-ray yield.

@@

We adopt the electron density profile as in Sect.~\ref{sec:simple} and model the temperature profile
of Coma with a constant central temperature of $kT= 8.25$~keV and a characteristic decline toward
the cluster periphery in accordance with a fit to the universal profile derived from cosmological
cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007} that also
follows the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{2005ApJ...628..655V}. This enables us to adopt the spatial and spectral distribution of CRs
according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution of
cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted lines). Additionally shown are integrals of the differential spectrum for
finited energy intervals across the angular apertures tested in this study (solid lines). These
modeled fluxes a compared to {\em Fermi} and VERITAS flux upper limits of the same energy intervals,
summarized in Table \ref{table:constraints}.  The gamma-ray flux limit of {\em Fermi} in the energy
interval of 1--3 GeV ($<0.4^\circ$) is most constraining but still above the model predictions that
assume an optimistic value of the shock acceleration efficiency and CR transport parameters as laid
out above.  In our further analysis, we use the most constraining {\em Fermi} flux limits in the
energy interval of 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the relative CR pressure, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$ as a function of radial distance, $R$, from the Coma cluster
center (red lines) and contained within $R$ (blue lines). All radii are shown in units of the virial
radius, $R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_p c$ as found in the cosmological simulations which reflects the high
cooling rates at low CR energies. The relative CR pressure rises toward the outer regions on account
of the higher efficiency of CR acceleration at the peripheral strong accretion shocks compared to
weak central flow shocks. Adiabatic compression of a mixture of CRs and thermal gas disfavours the
CR pressure relative to the thermal pressure on account of the softer equation of state of CRs. The
weak increase of $X_{\CR}$ toward the core is due to the comparably fast thermal cooling of gas
which diminishes the thermal support relative to that in CRs.

In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), those are a factor of 7.2 below the inferred upper limits of VERITAS (see Table
\ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches the simulated one at
energies $E_\gamma=200$ GeV). Hence we scale the averaged relative CR pressure of our model by that
factor, keep the spatial behavior and obtain an integrated limit of the relative CR pressure of
$X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit within the cluster virial radius of
$X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}).  This limit is less constraining by 50\% in
comparison to the simplified analytical model, which gives $X_\CR<0.1$. This difference is
explained by the concave curvature of the simulated spectrum which accumulates additional pressure
toward GeV energies in comparison to a pure power-law spectrum of $\alpha=2.3$.

Equivalently, the most constraining {\em Fermi} upper limit in the energy interval of 1--3 GeV
($<0.4^\circ$) is a factor of 1.1 larger than our model predictions (assuming $\alpha=2.1$ which is
very close to the simulated spectral index for the energy range of 1-3 GeV). Scaling our integrated
relative CR pressure profile yields the constraint of $X_{\CR}<0.018$ within 0.4$^{\circ}$ that
translates to a limit within the cluster virial radius of $X_\CR<0.024$ (dashed lines of Figure
\ref{fig:xcr}).  The $X_\CR$-constraint evaluated within the cluster virial radius is comparable to
the constraint in our simplified model of $X_\CR<0.025$.  Naturally, with the Fermi limits we probe
the region around GeV-energies that dominate the CR pressure and we do not expect any differences in
comparison to the simplified power-law model in comparison to the universal CR spectrum with its
concave curvature found in the simulations.

@@

\subsection{Minimum gamma-ray flux}
\label{sec:Fmin}
For clusters that host radio halos we are able to derive a minimum gamma-ray flux in the hadronic
model of radio halos -- where the radio-emitting electrons are secondaries from CR interactions. The
idea is based on the fact that a stationary distribution of CR electrons loses all its energy to
synchrotron radiation for strong magnetic fields ($B \gg B_{\rmn{CMB}} \simeq 3.2 \mu$G) so that the
ratio of gamma-ray to synchrotron flux becomes independent of the spatial distribution of CRs and
thermal gas \citep{article:Voelk:1989, article:Pohl:1994, article:Pfrommer:2008}, in particular with
$\alpha_{\nu}\simeq 1$ as the observed synchrotron spectral index.  Hence we can derive a minimum
gamma-ray flux in the hadronic model
\begin{equation}
\label{eq:Fmin}
\mathcal{F}_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\subsection{Constraining the Magnetic Field}
\label{sec:B}
We have seen in the previous section that we can obtain an absolute lower limit on the gamma-ray
emission in the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. Hence, we can
turn this around by using our upper limit on the gamma-ray emission (and by extension on the CR
pressure) to infer a lower limit on the magnetic field that is needed to explain the observed radio
emission \citep{article:PfrommerEnsslin:2004a}. This, however, assumes the validity of the hadronic
model of radio halos -- where the radio-emitting electrons are secondaries from CR
interactions. Lowering the gamma-ray constraint will tighten (increase) the magnetic field
limit. Once this gets in conflict with magnetic field measurements by means of other methods, e.g.,
Faraday rotation measure (RM)\footnote{Generally, Faraday RM analysis of the magnetic-field strength
  by, e.g., background sources through clusters are degenerate with the magnetic coherence scale and
  may be biased by the unknown correlation between magnetic and density fluctuations.}, this would
challenge the hadronic model of radio halos. The method that we use to constrain the magnetic field
inherits a dependence on the assumed spatial structure that we parametrize by
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magneto-hydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. In fact,
the magnetic field in the Coma cluster is among the best constrained, because its proximity permits
RM observations of seven radio-sources located at projected distances of 50 to 1500 kpc from the
cluster center. The best-fit model obtains $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and
$\alpha_{B} = 0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim for constraining the
central field strength, $B_{0}$, which depends on the magnetic decline, $\alpha_{B}$, that we vary
(within reasonable ranges of $\Delta\alpha_{B}=0.2$ as suggested from the Faraday RM studies) to
address this dependence. We proceed as follows:

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{{\em Figure TBC} Integral flux upper limits for Coma (this work, Tables~\ref{table:results}
  and \ref{table:fermi}) for different integration radii (shown with different colors), assuming
  $\alpha=2.1$. These are compared to integrated spectra of the same energy interval and aperture,
  assuming the universal gamma-ray spectrum of clusters \citep{article:PinzkePfrommer:2010}. To
  guide the eye, we show the underlying universal integral energy distribution of pion decay
  gamma-rays, $E_\gamma F_\gamma(>E_\gamma)$, resulting from hadronic interactions of CRs and ICM
  protons (dotted). Note that the {\em Fermi} limit for the energy interval of 1--3 GeV within the
  aperture of 0.4$^\circ$ is most constraining.}
\label{fig:spectrum}
\end{figure*}

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f5.eps}}
\end{center}
\caption{{\em Figure TBC} CR-to-thermal pressure, $X_{\mathrm{CR}}=\expval{P_{\CR}}/ \expval{P_{\mathrm{th}}}$, as a
function of radial distance from the center of the Coma. Model predictions are shown by dashed
lines \citep{article:PinzkePfrommer:2010} while these are scaled to match the most constraining
VERITAS upper limits within 0.2$^\circ$ (solid) and {\em Fermi} upper limits within 0.4$^\circ$
(dashed). We compare differential $X_\CR$-profiles (red) to integrated profiles
$X_{\CR}(<R/R_{\rmn{vir}})=\int_0^R P_\CR\, dV / \int_0^R P_{\rmn{th}}\, dV$ (blue) which we use to
compare to the upper limits.}
\label{fig:xcr}
\end{figure*}
